\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-11-24 sections/chameleon.tex
Fix latex errors
Update images
Assets and code listing

\begin{figure}[h]
    \centering
    \Description{Comparing Chameleon with \chameleon{}}
    \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

@@

\begin{lstlisting} [
    language=Haskell, caption={
    A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
}, label={listing:ifelse}
]
if a then b else c
a = "True"
\end{lstlisting}
For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example: \texttt{a} has type \texttt{Bool}, because \texttt{a} is the condition of an if statement; however \texttt{a} has type \texttt{String}, because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure}[h]
    \centering
    \Description{An anatomy of \chameleon{} interface}
    \includegraphics[width=\linewidth]{images/atonomy.png}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two windows: an editor window on the left and a
        debugging window on the right. The editor window is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains
        three panels. A message panel contains an error description (C), and, optionally,
        a list of candidate expression cards (D), a list of deduction steps (E), 
        and a control bar (F) to increment/decrement deduction step. 
        The conflicting types
        panel (G) shows two possible types inferred by the type inference
        engine. A relevant type information panel to show additional information
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \Description{Using \chameleon{} Type Compare Tool}
    \includegraphics[width=\linewidth]{images/intro-compare.pdf}
    \caption{
        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}. \chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type. In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
}
    \label{fig:compare}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{\linewidth}
        \centering
        \Description{Using \chameleon{} Candidate Expression Cards}
        \includegraphics[width=\linewidth]{images/intro-expression.pdf}
        \caption{
            \textbf{\chameleon{} with only candidate expression cards enabled.}
            \chameleon{} shows that the type error can
            occur in the definition of '\texttt{x}' or '\texttt{y}'.
        }
        \label{fig:expression}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\linewidth}
        \centering
        \Description{Using \chameleon{} Deduction Steps}
        \includegraphics[width=\linewidth]{images/intro-deductionstep.pdf}
        \caption{
            \textbf{\chameleon{} with both candidate expression cards and deduction steps
            enabled.}
            \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
        }
        \label{fig:deduction}
    \end{subfigure}
    \hfill
    \caption{ Candidate Expression Cards and Deduction Steps}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \Description{Showing an type error with a single candidate}
    \includegraphics[width=\linewidth]{images/single-error-location-example.png}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:single-candidate}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \Description{Chameleon Step}
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-F) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-E). To increment or decrement the deduction step can be intuitively thought of as to move the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

>>>>

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}

\begin{minted}{haskell}
if a then b else c
a = "True"
\end{minted}

For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example: \texttt{a} has type \texttt{Bool}, because \texttt{a} is the condition of an if statement; however \texttt{a} has type \texttt{String}, because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/atonomy.pdf}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two windows: an editor window on the left and a
        debugging window on the right. The editor window is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains
        three panels. A message panel contains an error description (C), and, optionally,
        a list of candidate expression cards (D), a list of deduction steps (E), 
        and a control bar (F) to increment/decrement deduction step. 
        The conflicting types
        panel (G) shows two possible types inferred by the type inference
        engine. A relevant type information panel to show additional information
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/intro-compare.pdf}
    \caption{
        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}. \chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type. In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
}
    \label{fig:compare}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/intro-candidate.pdf}
    \caption{
        \textbf{\chameleon{} with only candidate expression cards enabled.}
        \chameleon{} shows that the type error can
        occur in the definition of '\texttt{x}' or '\texttt{y}'.
    }
    \label{fig:expression}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/intro-deduction.pdf}
    \caption{
        \textbf{\chameleon{} with both candidate expression cards and deduction steps
        enabled.}
        \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
    }
    \label{fig:deduction}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/single-error-location-example.png}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:single-candidate}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-F) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-E). To increment or decrement the deduction step can be intuitively thought of as to move the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}
